Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Add genomics products refs

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets, 
and has been instrumental in the delivery of datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
it is complex, vertically integrated software
with custom storage formats that are not intended for reuse.
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

>>>>

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets, 
and has been instrumental in the delivery of datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
it is complex, vertically integrated software 
(i.e., consisting of tightly coupled components)
with custom storage formats that are not intended for reuse.
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
awshealthomics2024,microsoftgenomics2024}
or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.
